# The R-INLA package

**Learning objectives:**

- Learning about the `inla()` function 
    - basics argument  
    - default set up and how to change it
    - interacting with its output
    
- Useful functions in R-INLA    

## Linear predictor (LP)

```{r}
library(INLA)
library(ggplot2)
```


1. step one: writing linear predictor as an R formula 

$$Response \quad variable \sim fixed + random \quad effect$$

Random effect are put inside `f()` with a name of model arguments ("iid")


$$\eta_i = \beta_o + \beta_1x_1 + \beta_2x_2 + u_i $$

$$u_i \sim N(0, \sigma^2_u) $$

```{r eval=FALSE} 
y ~ x1 + x2 + f(i, model = "iid")
```


```{r eval=FALSE} 
y ~ 0 = b0 + x1 + x2 + f(i, model = "iid")
```

## Inla()

Keys argument: 

- `formula`

- `data`

- `family` : type of likelihhod distributon we want to use as model

```{r}
(inla.models()$likelihood)$poisson
#inla.doc("poisson") # replace poisson with other family
```

- `control.compute` : take a list with "key = value" of other variable that should be computed. Example with deviance information criterion (DIC)

- `control.predictor` : take a list that specify predictor variable (`link`) but also if marginal densities of LP should be computed


## Priors specification

```{r}
(inla.models()$prior$normal)
inla.doc("Normal")
```


You should always check the doc (or do yourself some cheat sheet) because like every software they could use other convention : use of precision instead of variance in $N()$. 

INLA use some default setting but you can overwrite at:

* the "default level":  `inla.set.control.fixed.default()`

* Inside `inla()`: `control.fixed = prior.fixed` prior.fixed is a list

The hyperparameters are specified in a list ("key=value") with the `hyper` arguments inside the `f()`: 

- `inital`: where to start 

- `prior`: prior distribution

- `param`: vector of the parameters for distribution

- `fixed`: fixed value or not


You can use Penalized Complexity or PC priors. 


## Example: mortality rates following surgery

This is an illustration of what we seens previously but it also will add few important functions to help us analyze output of `inla()`.

```{r}
Surg
```

Here we want to investigate mortality rate (r)

### Model

$$Y_i \sim Binom(n_i, p_i), i = 1, .., 12$$

$$logit(pi) = \alpha + u_i, u_i \sim N(o, \sigma^2) $$

Reminder: logit is the log odds ($ln(\frac{p_i}{1 -p_i}$)

Default prior for $\alpha \sim N(0, \frac{1}{\tau})$

We will PC for $\u_i$ with 

$P(\sigma > 1) = 0.01$

Define prior as a list:

```{r}
prior.prec <- list(prec = list(prior = "pc.prec",
                               param = c(1, 0.01)))
```

Define formula object:

```{r}
formula <- r ~ f(hospital, model = "iid", hyper = prior.prec)
```

Calling inla() :

```{r}
res <- inla(formula,
  data = Surg,
  family = "binomial", Ntrials = n,
  control.predictor = list(compute = TRUE), # marginal densities of LP
  control.compute = list(dic = TRUE) # DIC 
)
```

### resuts!

```{r}
summary(res)
```

You can access specific part of the result using `$`

```{r}
res$summary.fixed
```

We asked `compute = TRUE` so we get : 

- `summary.linear.predictor` : **df** with linear predictors

- `summary.fitted.values` : **df** with fitted values obtained by transforming the linear predictors by the inverse of the link function (if like me you have trouble with log values)

- `marginals.linear.predictor`: **list** with the posterior marginals of the linear predictors

- `marginals.fitted.values`: **list** with the posterior marginals of the fitted values obtained by transforming the linear predictors by the inverse of the link function

You can acces them like that:

```{r}
res$summary.fitted.values
```

Or using r-inla functions to manipulate them: 

- `inla.L.marginal()` L is for "Letters":

    * e for expectations 
    * q for quantiles  
    * s for spline smoothings  
    * t for transform (if you want varianec instead of precision)  
    * z for summary statistics


```{r}
library(ggplot2)
alpha <- res$marginals.fixed[[1]]
ggplot(data.frame(inla.smarginal(alpha)), aes(x, y)) +
  geom_line() +
  theme_bw()
```

```{r}
marg.variance <- inla.tmarginal(function(x) 1/x,
res$marginals.hyperpar$"Precision for hospital")

ggplot(data.frame(inla.smarginal(marg.variance)), aes(x, y)) +
  geom_line() +
  theme_bw()
```

```{r, eval=FALSE}
list_marginals <- res$marginals.fitted.values

marginals <- data.frame(do.call(rbind, list_marginals))
marginals$hospital <- rep(names(list_marginals),
                           times = sapply(list_marginals, nrow))

library(ggplot2)
ggplot(marginals, aes(x = x, y = y)) + geom_line() +
  facet_wrap(~ hospital) +
  labs(x = "", y = "Density") +
  geom_vline(xintercept = 0.1, col = "gray") +
  theme_bw()
```

## Control variables to compute approximations

`control.inla` = list(int.strategy = "grid")

## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>
<summary> Meeting chat log </summary>

```
LOG
```
</details>
