# Areal data

**Learning objectives:**

-   Learn how to work with **Aereal data**
-   How to compute neighborhood matrices, expected counts, and SIRs.
-   How to fit spatial and spatio-temporal disease risk models using the R-INLA package

## Introduction

This chapter is about Aereal data. We will be looking at some real world examples provided by the {SpatialEpi} package, and do some math for measuring disease risks before building the model.

Examples of **Areal data** are:

-   the number of cancer cases in counties
-   the number of road accidents in provinces
-   the proportion of people living in poverty in census tracts.

**Measure of disease risk in areas**:

-   standardized incidence ratio (SIR) which is defined as the ratio of the observed to the expected counts
-   Bayesian hierarchical models

**Spatial model: Besag-York-Mollié (BYM) model**

-   autocorrelation in temporal and spatial structure
-   spatial random effects

**Data used**: lung cancer in Pennsylvania counties, USA, from the {SpatialEpi} package

## Spatial neighborhood matrices

Packages needed:

```{r 05-Aereal-data-1, message=FALSE, warning=FALSE, paged.print=FALSE}
library(tidyverse)
library(sf)
library(spdep)
library(SpatialEpi)
library(INLA)
library(cowplot)
```

A **spatial model** is a particular type of model as it takes consideration of autocorrelation between spatial points and so some *weights* are considered to identify the points' proximity.

$$y=\sum_{i=1}^n{w_if(x_i)}$$

$\mathit{W}$ is the **proximity matrix** or the **inverse distance between areas** and spatially connects areas $i$ and $j$, its elements $(i,j)$ , are:

$$i=(1,...,n)$$ $$j=(1,...,m)$$

$$W=\begin{bmatrix}
(1,1) & (1,2) & ... & (1,j)\\ 
(2,1) & ... & ... & (2,j)\\ 
... & ... & ... & ...\\ 
(i,1) & ...& ... &(i,j)\\
\end{bmatrix}$$


The most simple cases are:

$$w_{i,j}=\left\{\begin{matrix}
1 & \text{ if i and j share same boundaries}\\
0 & \text{ otherwise}
\end{matrix}\right.$$


As an alternative, a **standardized matrix** can be used, adjusted for the total number of neighbors in each area:

$$w_{std,i,j}=\frac{w_{i,j}}{\sum_{j=1}^n{w_{ij}}}$$


Let's have a look at the data: **Lung Cancer in Pennsylvania** 
```{r 05-Aereal-data-2}
# install.packages("SpatialEpi")
# library(SpatialEpi)

map <- pennLC$spatial.polygon
plot(map)
```

And construct the neighbors list for each county:

```{r 05-Aereal-data-3,comment=""}
# install.packages("spdep")
# library(spdep)
nb <- spdep::poly2nb(map)
nb%>%class
head(nb,3)
```



```{r 05-Aereal-data-4, comment=""}
d <- data.frame(county = names(map), 
                neigh = rep(0, length(map)))

rownames(d) <- names(map)

map <- SpatialPolygonsDataFrame(map, d, match.ID = TRUE)
map$neigh[nb[[2]]] <- 1
map$neigh[nb[[44]]] <- 1
map$neigh[nb[[58]]] <- 1
map%>%as_tibble()%>%head
```

```{r 05-Aereal-data-5}
coord <- coordinates(map)
map$long <- coord[, 1]
map$lat <- coord[, 2]
map$ID <- 1:dim(map@data)[1]
map@proj4string
map%>%as_tibble()%>%head
```

```{r 05-Aereal-data-6}
# install.packages("sf")
# library(sf)
mapsf <- sf::st_as_sf(map)
```

```{r 05-Aereal-data-7, fig.cap="Neighbors of areas 2, 44 and 58 of Pennsylvania"}
# ggplot(mapsf)+
#   geom_sf()

ggplot(mapsf) + 
  geom_sf(aes(fill = as.factor(neigh))) +
  geom_text(aes(long, lat, label = ID), color = "white") +
  guides(fill = "none") +
  theme_bw() 
```


## Standardized incidence ratio

Estimation of the **disease risk estimates in each of the areas**


$$SIR_i=\frac{Y_i}{E_i}$$ $E_i$ is the **expected counts** or the expected total number of cases, is the sum of the multiplication of the rate of the number of cases divided by population in $j$  $r_j^{(s)}$, and $n_j^{(i)}$ the population in stratum $j$ of area $i$.

$$E_i=\sum_{j=1}^m{r_{j}^{(s)}n_{j}^{(i)}}$$

$$SIR_i=\begin{Bmatrix}
>1 & \text{higher} \\ 
=1 & \text{equal} \\
<1 & \text{lower} 
\end{Bmatrix} \text{risk than expected in area i}$$

When applied to mortality data, the ratio is known as the **standardized mortality ratio (SMR)**.

Example:

```{r 05-Aereal-data-8}
d <- pennLC$data%>%
  group_by(county) %>% 
  summarize(Y = sum(cases))

head(d)
```

```{r 05-Aereal-data-9}
pennLC$data <- pennLC$data %>%
  arrange(county,race,gender,age)
```

```{r 05-Aereal-data-10}
E <- expected(
  population = pennLC$data$population,
  cases = pennLC$data$cases, n.strata = 16
)
```


```{r 05-Aereal-data-11}
d$E <- E[match(d$county, unique(pennLC$data$county))]


head(d)
```

```{r 05-Aereal-data-12}
d$SIR <- d$Y / d$E
```

```{r 05-Aereal-data-13}
head(d)
```

```{r 05-Aereal-data-14}
map <- merge(map, d)
```

```{r 05-Aereal-data-15}
mapsf <- st_as_sf(map)
```

```{r 05-Aereal-data-16, fig.cap="SIR of lung cancer in Pennsylvania counties"}
ggplot(mapsf) + geom_sf(aes(fill = SIR)) +
  scale_fill_gradient2(
    midpoint = 1, low = "blue", mid = "white", high = "red"
  ) +
  theme_bw()
```

## Spatial small area disease risk estimation

$$Y\sim{Po(E_{i} \theta_{i})}, i=1,..,n,$$ $$log(\theta_i)=\alpha+u_i+v_i$$

$\alpha$ is the overall risk, $u_i$ the random effect in area i, and $v_i$ is some noise $v_i\sim{N(0,\sigma_v^{2})}$

To include covariates to quantify risk factors and other random effects:

$\alpha=\mathbf{d_i \beta}$

The **Besag-York-Mollié (BYM) model** (Besag, York, and Mollié 1991) $u_i$, which is the spatial random effect, has a **Conditional Autoregressive (CAR)** distribution:

$$u_i|\mathbf{u_{-i}}\sim{N(\bar{u_{\sigma_i},\frac{\sigma_u^2}{n_{\sigma_i}}})}$$

$\bar{u}_{\sigma_i}=n_{\sigma_i}^{-1}\sum_{j\epsilon\sigma_i}{u_j}$

In R-INLA, the formula of the BYM model is specified as follows:

$$Y\sim{f_1(besag)+f_2(iid)}$$

```{r 05-Aereal-data-17}
formula <- Y ~
  f(idareau, model = "besag", # CAR distribution
    graph = g,                # neighborhood structure
    scale.model = TRUE) +
  f(idareav, model = "iid")
```

(The BYM model can also be specified with the model "bym")

### Spatial modeling of lung cancer in Pennsylvania

To calculate the relative risks of lung cancer in the Pennsylvania counties, a new parametrization of the BYM model is called BYM2, **bym2** model, it defines the components as:

$$\mathbf{b}=\frac{1}{\sqrt{\pi_b}(\sqrt{1-\phi}\mathit{v}_*+\sqrt{\phi}\mathit{u_*})}$$

$\pi_b$ is the precision parameter and controls the marginal variance of the weighted sum of $\mathit{u_*}$ and $\mathit{v_*}$, $\phi$ is between 0 and 1, and measures the proportion of the marginal variance explained by $\mathit{u_*}$

$$\text{BYM2}=\left\{\begin{matrix}
\phi=1 & \text{one spatial model} \\ 
 \phi=0 & \text{one unstructured spatial noise}
\end{matrix}\right.$$

$$Y\sim{f(.)}$$

```{r 05-Aereal-data-18}
formula <- Y ~ f(idarea, # is the index variable of the area
                 model = "bym2", 
                 graph = g)
```

```{r 05-Aereal-data-19}
map$idarea <- 1:nrow(map@data)
```

As a **rule of thumb**, given the PC prior for the marginal precision $\pi_b$:

$$P((1/\sqrt{\pi_b})>U)=\alpha$$

we use: $$U=0.5/0.31$$ and $$\alpha=0.01$$

```{r 05-Aereal-data-20}
prior <- list(
  prec = list(
    prior = "pc.prec",
    param = c(0.5 / 0.31, 0.01)),
  phi = list(
    prior = "pc",
    param = c(0.5, 2 / 3))
  )
```

```{r 05-Aereal-data-21}

nb <- poly2nb(map)
head(nb)
```

```{r 05-Aereal-data-22}
spdep::nb2INLA("map.adj", nb)

g <- INLA::inla.read.graph(filename = "map.adj")
```

```{r 05-Aereal-data-23}
# it requires BiocManager::install("Rgraphviz")
plot(g)
```

```{r 05-Aereal-data-24}
formula <- Y ~ f(idarea, 
                 model = "bym2", 
                 graph = g, 
                 hyper = prior)
```

And fit the model:

```{r 05-Aereal-data-25, message=FALSE, warning=FALSE, paged.print=FALSE}
res <- inla(formula,
  family = "poisson", 
  data = map@data,
  E = E, 
  control.predictor = list(compute = TRUE)
)
```

```{r 05-Aereal-data-26}
summary(res)
```

```{r 05-Aereal-data-27}
head(res$summary.fitted.values)
```

```{r 05-Aereal-data-28}
map$RR <- res$summary.fitted.values[, "mean"]
map$LL <- res$summary.fitted.values[, "0.025quant"]
map$UL <- res$summary.fitted.values[, "0.975quant"]

summary(map@data[, c("RR", "LL", "UL")])
```

```{r 05-Aereal-data-29}
mapsf <- st_as_sf(map)

gRR <- ggplot(mapsf) + geom_sf(aes(fill = RR)) +
  scale_fill_gradient2(
    midpoint = 1, low = "blue", mid = "white", high = "red",
    limits = c(0.7, 1.5)
  ) +
  theme_bw()
```

```{r 05-Aereal-data-30}
gLL <- ggplot(mapsf) + geom_sf(aes(fill = LL)) +
  scale_fill_gradient2(
    midpoint = 1, low = "blue", mid = "white", high = "red",
    limits = c(0.7, 1.5)
  ) +
  theme_bw()
```

```{r 05-Aereal-data-31}
gUL <- ggplot(mapsf) + geom_sf(aes(fill = UL)) +
  scale_fill_gradient2(
    midpoint = 1, low = "blue", mid = "white", high = "red",
    limits = c(0.7, 1.5)
  ) +
  theme_bw()
```

```{r 05-Aereal-data-32}
cowplot::plot_grid(gRR, gLL, gUL, ncol = 1)
```

The posterior mean of the BYm2 random effect $\mathbf{b}$:

```{r 05-Aereal-data-33}
mapsf$re <- res$summary.random$idarea[1:67, "mean"]

ggplot(mapsf) + geom_sf(aes(fill = re)) +
  scale_fill_gradient2(
    midpoint = 0, low = "blue", mid = "white", high = "red"
  ) +
  theme_bw()
```

## Spatio-temporal small area disease risk estimation

$$Y_{ij}\sim{Po(E_{ij}\phi_{ij})}$$

A **Spatio-temporal model with parametric time trends**

$$log(\phi_{ij}=\alpha+\mathit{u_i}+\mathit{v_i}+(\beta+\delta_i) \times \mathit{t_j}$$

$$Y\sim f(bym)+f(iid)+ \text{idtime}$$

```{r 05-Aereal-data-34}
formula <- Y ~ 
  f(idarea,    # area index
    model = "bym", 
    graph = g) + 
  f(idarea1,   # area index
    idtime,   
    model = "iid") + 
  idtime.      # time index
```

-   `f(idarea,model = "bym", graph = g)` is the area random effect $\mathit{u_i}+\mathit{v_i}$.

-   `f(idarea1, idtime, model = "iid")`. is the differential time trend $\delta_i \times \mathit{t}_j$

-   `idtime` is the global trend $\beta \times \mathit{t}_j$

An alternative is from **Knorr-Held (2000)** specify models with interaction between space and time:

```{r 05-Aereal-data-35}
formula <- Y ~ f(idarea, model = "bym", graph = g) +
  f(idtime, model = "rw2") +
  f(idtime1, model = "iid") +
  f(idareatime, model = "iid")
```

```{r 05-Aereal-data-36}
# f(idareatime, model = "iid")
```

```{r 05-Aereal-data-37, eval=FALSE}
f(idtime,
  model = "iid",
  group = idarea, 
  ontrol.group = list(model = "besag", graph = g)
)
```

```{r 05-Aereal-data-38, eval=FALSE}
f(idarea,
  model = "iid",
  group = idtime, control.group = list(model = "rw2")
)
```

```{r 05-Aereal-data-39, eval=FALSE}
f(idarea,
  model = "besag", graph = g,
  group = idtime, control.group = list(model = "rw2")
)
```

## Conclusions

Things to consider:

-   Misaligned Data Problem (MIDP)
-   Modifiable Areal Unit Problem (MAUP)
    -   ecological inference problem:
        -   aggregation bias
        -   specification bias

## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/AqCiwVMEwfc")`

<details><summary>Meeting chat log</summary>
```
01:07:44	defuneste:	https://r-spatial.github.io/spdep/articles/nb.html
01:08:16	Oluwafemi Oyedele:	Thank you !!!
```
</details>

`r knitr::include_url("https://www.youtube.com/embed/k7U0ufIqsys")`

<details><summary>Meeting chat log</summary>
```
00:02:40	Oluwafemi Oyedele:	Hi, Good Evening !!!
00:20:29	Oluwafemi Oyedele:	Can you share the link ?
00:21:06	Olivier Leroy:	https://stats.stackexchange.com/questions/120964/fixed-effect-vs-random-effect-when-all-possibilities-are-included-in-a-mixed-eff
00:57:24	Federica Gazzelloni:	pennLC$data$cases[pennLC$data$county=="adams"]
```
</details>
