# Spatial modeling of areal data. Lip cancer in Scotland

**Learning objectives:**

- Explore sp/sf data abstraction 

- Practice spatial modelling for aerial data (SIR and BYM)

- Display result in tables, plots (ggplot2) and interactive map (leaflet)

## Data set: wrangling 

```{r loading lib}
library(sp)
library(SpatialEpi)
library(sf)
library(leaflet)
```

The data set cover Scoltland counties:

- Observed and expected cases of lip cancer in males

- % of opop in Agriculture, Fishing, Forestry (AFF)


### Map of Scotland counties

We will be using both `{sp}` and `{sf}` just for the sake of understanding: 

```{r lodaing_ploting_data}
data(scotland)
map <- scotland$spatial.polygon
plot(map)
```

The workflow in `{sp}` an start with getting all EPSG codes and proj4 fron `rgdal`:  

```{r sp_proj}
# sp
codes <- rgdal::make_EPSG()
codes[which(codes$code == "27700"), ]
codes[codes$code == "4326",]
# saving it for other purposes
map_3 <- map_2 <-map
# change in unit was done by "hand"
proj4string(map) <- "+proj=tmerc +lat_0=49 +lon_0=-2
+k=0.9996012717 +x_0=400000 +y_0=-100000 +datum=OSGB36
+units=km +no_defs"
```

An other option if you want to stay in `{sp}` but avoid `{rgdal}`. You have to keep in mind that `{sp}`  use S4 classes. 

```{r mix_sp_sf_proj}
# partially sf
map_2 <- map
(proj4string(map_2) <- sp::CRS(SRS_string = sf::st_crs(27700)$wkt))
```

Doing it with full `{sf}`:

```{r full_sf}
# More sf 
map_3 <- st_as_sf(map)
map_3 <- st_set_crs(map_3, 27700) # print st_set_crs()
map_3$name <- sapply(map@polygons, function(x) x@ID)
# an other way of doing it provided by Paula Moraga
# sapply(slot(map, "polygons"), function(x){slot(x, "ID")})
# I did not know about slot()
```

More on proj4string and **crs** as **wkt** here: https://r-spatial.org/r/2020/03/17/wkt.html

For leaflet we will need to transform the CRS:

```{r}
# require rgdal
map <- sp::spTransform(map,
                   CRS("+proj=longlat +datum=WGS84 +no_defs"))

map_3 <- sf::st_transform(map_3, 4326)
```

### Data

With `{sp}`:

```{r adding_data}
d <- scotland$data[,c("county.names", "cases", "expected", "AFF")]
names(d) <- c("county", "Y", "E", "AFF")
d$SIR <- d$Y / d$E
rownames(d) <- d$county
map <- SpatialPolygonsDataFrame(map, d, match.ID = TRUE)
```


With `{sf}`:

```{r}
# you can also use dplyr verbs
map_3m <- merge(map_3, d, by.x = "name", by.y = "county")
```

## Mapping SIRs

A first version:

```{r first_leaflet}
# a base map, tile by default are OSM
l <- leaflet(map) |> addTiles()

pal <- leaflet::colorNumeric(palette = "YlOrRd", domain = map$SIR)

l |>
  # border
  addPolygons(
    color = "grey", weight = 1,
    fillColor = ~ pal(SIR), fillOpacity = 0.5
  ) |>
  # SIR
  addLegend(
    pal = pal, values = ~SIR, opacity = 0.5,
    title = "SIR", position = "bottomright"
  )
```

Building on it: 

- adding good label:  
  * Use `stringf()`   
  * use of `htmltools::HTML`

```{r labels}
labels <- sprintf("<strong> %s </strong> <br/>
  Observed: %s <br/> Expected: %s <br/>
  AFF: %s <br/> SIR: %s",
  map$county, map$Y, round(map$E, 2),
  map$AFF, round(map$SIR, 2)
) |>
  lapply(htmltools::HTML)
```


Then we add it to our previous codes: 

```{r addinglabels}
l |>
  addPolygons(
    color = "grey", weight = 1,
    fillColor = ~ pal(SIR), fillOpacity = 0.5,
    highlightOptions = highlightOptions(weight = 4),
    ## here
    label = labels,
    labelOptions = labelOptions(
      style = list(
        "font-weight" = "normal",
        padding = "3px 8px"
      ),
      textsize = "15px", direction = "auto"
    )
  ) |>
  addLegend(
    pal = pal, values = ~SIR, opacity = 0.5,
    title = "SIR", position = "bottomright"
  )
```

SIR can be problematic if we have low population -> Modeling!

## Modeling

### Our model:

First assumptions: Observed count is Poisson distributed ($y_i$)

$$Y_i \sim Poisson(E_i\theta_i), i = 1, ..., n$$

With $E_i$ is the expected count and $\theta_i$ is the relative risk in area $i$.

$$log(\theta_i) = \beta_0  + \beta_1 \times AFF_i + u_i + v_i$$


$$\mu | \mu_{-i} \sim N (\bar{u_{\delta_i}}, \frac{\sigma^2_u}{n_{\delta_i}})$$

$$v_i \sim N(0, \sigma²_v)$$

### neighborhood matrix

```{r nb}
library(spdep)
library(INLA)
nb <- poly2nb(map) # same result with map_3
nb2INLA("map.adj", nb)
g <- inla.read.graph(filename = "map.adj")
```


### Using INLA

With have two random effects $u$ and $v$:

```{r init_id}
map$idareau <- 1:nrow(map@data) # u
map$idareav <- 1:nrow(map@data) # v
# sf map_3iareau <- 1:nrow(map_3)
```

Same model than chapter 5:

```{r formula}
formula <- Y ~ AFF +
  f(idareau, model = "besag", graph = g, scale.model = TRUE) +
  f(idareav, model = "iid")
```

```{r  inla}
res <- inla(formula,
  family = "poisson", data = map@data,
  E = E, control.predictor = list(compute = TRUE),
  # this need to be added for latter!!!
  #control.compute=list(return.marginals.predictor=TRUE)

)
summary(res)
```
Greatly surprise by:

- this was quick!  

- AFF increases lip cancer risk (and it is clear)

```{r ggplotRisk}
library(ggplot2)
marginal <- inla.smarginal(res$marginals.fixed$AFF) # list
marginal <- data.frame(marginal)
ggplot(marginal, aes(x = x, y = y)) + geom_line() +
  labs(x = expression(beta[1]), y = "Density") +
  geom_vline(xintercept = 0, col = "black") + theme_bw()
```

## Mapping relative risks

We have mean and intervals how do we represent that? (labels)

1. Add mean and intervals to data 

```{r interval_RR}
map$RR <- res$summary.fitted.values[, "mean"]
map$LL <- res$summary.fitted.values[, "0.025quant"]
map$UL <- res$summary.fitted.values[, "0.975quant"]
```

2. add it to the labels

(RR: relative risk)


```{r}
pal <- colorNumeric(palette = "YlOrRd", domain = map$RR)

labels <- sprintf("<strong> %s </strong> <br/>
  Observed: %s <br/> Expected: %s <br/>
  AFF: %s <br/> SIR: %s <br/> RR: %s (%s, %s)",
  map$county, map$Y, round(map$E, 2),
  map$AFF, round(map$SIR, 2), round(map$RR, 2),
  round(map$LL, 2), round(map$UL, 2)
) %>% lapply(htmltools::HTML)

lRR <- leaflet(map) %>%
  addTiles() %>%
  addPolygons(
    color = "grey", weight = 1, fillColor = ~ pal(RR),
    fillOpacity = 0.5,
    highlightOptions = highlightOptions(weight = 4),
    label = labels,
    labelOptions = labelOptions(
      style =
        list(
          "font-weight" = "normal",
          padding = "3px 8px"
        ),
      textsize = "15px", direction = "auto"
    )
  ) %>%
  addLegend(
    pal = pal, values = ~RR, opacity = 0.5, title = "RR",
    position = "bottomright"
  )

lRR

```


## Exceedance probabilities

Probabilities of RR ($\theta_i$) estimates being greater than a given threshold value ($c$).

$$P(\theta_i > c) = 1 - P(\theta_i \leq c) $$

```{r}
# 1 - inla.pmarginal(q = c, marginal = marg)
res <- readRDS("data/res_ch6")
c = 2
marg <- res$marginals.fitted.values[[1]] # first county in DF
1 - inla.pmarginal(q = c, marginal = marg)
```

We want it for every counties: 

```{r}
exc <- sapply(res$marginals.fitted.values,
FUN = function(marg){1 - inla.pmarginal(q = 2, marginal = marg)})

map$exc <- exc
```


Then we add it to our labels: 

```{r}
pal <- colorNumeric(palette = "YlOrRd", domain = map$exc)

labels <- sprintf("<strong> %s </strong> <br/>
  Observed: %s <br/> Expected: %s <br/>
  AFF: %s <br/> SIR: %s <br/> RR: %s (%s, %s) <br/> P(RR>2): %s",
  map$county, map$Y, round(map$E, 2),
  map$AFF, round(map$SIR, 2), round(map$RR, 2),
  round(map$LL, 2), round(map$UL, 2), round(map$exc, 2) ## <- here
) %>% lapply(htmltools::HTML)

lexc <- leaflet(map) %>%
  addTiles() %>%
  addPolygons(
    color = "grey", weight = 1, fillColor = ~ pal(exc),
    fillOpacity = 0.5,
    highlightOptions = highlightOptions(weight = 4),
    label = labels,
    labelOptions = labelOptions(
      style =
        list(
          "font-weight" = "normal",
          padding = "3px 8px"
        ),
      textsize = "15px", direction = "auto"
    )
  ) %>%
  # in lengend instead of RR
  addLegend(
    pal = pal, values = ~exc, opacity = 0.5, title = "P(RR>2)",
    position = "bottomright"
  )
lexc
```



## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>
<summary> Meeting chat log </summary>

```
LOG
```
</details>
