# Spatio-temporal modeling of areal data. Lung cancer in Ohio

**Learning objectives:**


Risk of lung cancer in Ohio, USA, from 1968 to 1988 using the R-INLA package:

- how to calculate the expected disease counts using indirect standardization, and the standardized incidence ratios (SIRs).

- how to fit a Bayesian spatio-temporal model to obtain disease risk estimates for each of the Ohio counties and years of study.

- how to create static and interactive maps and time plots of the SIRs and disease risk estimates.


## Data

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# remotes::install_github("Paula-Moraga/SpatialEpiApp")
library(SpatialEpiApp)
namecsv <- "SpatialEpiApp/data/Ohio/dataohiocomplete.csv"
dohio <- read.csv(system.file(namecsv, package = "SpatialEpiApp"))
head(dohio)
```

Map of Ohio

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
library(rgdal)
library(sf)

nameshp <- system.file(
"SpatialEpiApp/data/Ohio/fe_2007_39_county/fe_2007_39_county.shp",
package = "SpatialEpiApp")
map <- readOGR(nameshp, verbose = FALSE)

plot(map)
```


Calculate the observed and expected counts, and the SIRs for each county and year, and create a data frame.



```{r message=FALSE, warning=FALSE, paged.print=FALSE}
library(tidyverse)
d <- dohio %>%
  group_by(county=NAME,year) %>%
  summarize(Y=sum(y))%>%
  ungroup()%>%
  arrange(year)
head(d)
```

## Expected cases

- indirected standardization
- strata: 2 races X 2 genders = 2X2 = 4
```{r}
library(SpatialEpi)
n.strata = 4
dohio <- dohio%>%
  arrange(county,year,gender,race)
  
E <- expected(population = dohio$n,
              cases = dohio$y,
              n.strata = n.strata)
```


```{r}
nyears <- length(unique(dohio$year))
countiesE <- rep(unique(dohio$NAME),
                 each = nyears)
```

```{r}
ncounties <- length(unique(dohio$NAME))
yearsE <- rep(unique(dohio$year),
              times = ncounties)

dE <- data.frame(county = countiesE, year = yearsE, E = E)

head(dE)
```


```{r}
d <- merge(d, dE, by = c("county", "year"))
head(d)
```

## SIRs


$$SIR_{ij}=\frac{Y_{ij}}{E_{ij}}$$

```{r}
d$SIR <- d$Y / d$E
head(d)
```

```{r}
dw <- d%>%
  pivot_wider(names_from=year,values_from = c("Y","E","SIR"))
head(dw)
```


## Mapping


```{r}
map@bbox
map@data[1:2,]

map <- merge(map, dw, by.x = "NAME", by.y = "county")
```



```{r}
map_sf <- st_as_sf(map)
```


```{r}
map_sf <- gather(map_sf, year, SIR, paste0("SIR_", 1968:1988))
```

```{r}
map_sf$year <- as.integer(substring(map_sf$year, 5, 8))
```

```{r}
ggplot(map_sf) + 
  geom_sf(aes(fill = SIR)) +
  facet_wrap(~year, dir = "h", ncol = 7) +
  ggtitle("SIR") + theme_bw() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank()
  ) +
  scale_fill_gradient2(
    midpoint = 1, low = "blue", mid = "white", high = "red"
  )
```


### Time plots of SIRs

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
g <- ggplot(d, 
            aes(x = year, y = SIR, 
                   group = county, 
                   color = county)) +
  geom_line() + 
  geom_point(size = 2) + 
  theme_bw()+
  theme(legend.position = "none")

library(gghighlight)
g + gghighlight(county == "Lorain")
```

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
library(plotly)
ggplotly(g)
```

## Modeling

Relative risk of lung cancer for each Ohio county and year is estimated with the **Bernardinelli model (Bernardinelli et al. 1995)**.

$$Y_{ij}\sim Po(E_{ij}\theta_{ij})$$
$$log(\theta_{ij})=\alpha+u_i+v_i+(\beta+\delta_i) \text{ x }t_j$$

### Neighborhood matrix

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
library(INLA)
library(spdep)
map%>%class
nb <- poly2nb(map)
head(nb)
```
```{r eval=FALSE}
?poly2nb
?nb2INLA
?inla.read.graph
```


```{r}
nb2INLA("map.adj", nb)
g <- inla.read.graph(filename = "map.adj")
```

### Inference using INLA

```{r}
d$idarea <- as.numeric(as.factor(d$county))
d$idarea1 <- d$idarea
d$idtime <- 1 + d$year - min(d$year)
```


And the model:
```{r}
formula <- Y ~ f(idarea, model = "bym", graph = g) + f(idarea1, idtime, model = "iid") + idtime
```

```{r}
res <- inla(formula,
  family = "poisson", 
  data = d, 
  E = E,
  control.predictor = list(compute = TRUE)
)
```


## Mapping relative risks

```{r}
d$RR <- res$summary.fitted.values[, "mean"]
d$LL <- res$summary.fitted.values[, "0.025quant"]
d$UL <- res$summary.fitted.values[, "0.975quant"]
```



```{r}
map_sf <- merge(
  map_sf, d,
  by.x = c("NAME", "year"),
  by.y = c("county", "year")
)
```



```{r}
ggplot(map_sf) + 
  geom_sf(aes(fill = RR)) +
  facet_wrap(~year, dir = "h", ncol = 7) +
  ggtitle("RR") + theme_bw() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank()
  ) +
  scale_fill_gradient2(
    midpoint = 1, low = "blue", mid = "white", high = "red"
  )
```

Animate the plot:

```{r eval=FALSE}
library(gganimate)
ggplot(map_sf) + 
  geom_sf(aes(fill = RR)) +
  theme_bw() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank()
  ) +
  scale_fill_gradient2(
    midpoint = 1, low = "blue", mid = "white", high = "red"
  ) +
  transition_time(year) +
  labs(title = "Year: {round(frame_time, 0)}")
```


To save the animation:
```{r}
# anim_save("oh_map.gif")
```


## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/-3mKJn8AMZI")`

<details>
<summary> Meeting chat log </summary>

```
00:03:08	Olivier Leroy:	https://grass.osgeo.org/
00:03:49	Olivier Leroy:	https://cran.r-project.org/web/packages/rgrass/index.html
00:21:21	Olivier Leroy:	https://www.ngs.noaa.gov/datums/horizontal/north-american-datum-1983.shtml
00:21:24	Olivier Leroy:	a datum
00:21:46	Olivier Leroy:	ND stands for North American
00:21:52	Olivier Leroy:	Datum
00:27:13	Olivier Leroy:	can you try "Lorain" ?
00:54:21	Olivier Leroy:	https://paulamoraga.shinyapps.io/spatialepiapp/
```
</details>
