# Spatial modeling of geostatistical data. Malaria in The Gambia

**Learning objectives:**

-   How to **fit a geostatistical model to predict malaria prevalence in The Gambia** using the **stochastic partial differential equation (SPDE)** approach and the R-INLA package
-   How to create a **triangulated mesh** that covers The Gambia, the projection matrix and the data stacks to fit the model
-   How to **manipulate the results to obtain the malaria prevalence** predictions, and 95% credible intervals denoting uncertainty.

## Data & data preparation

### data

```{r 09-Spatial-modeling-of-geostatistical-data-1, message=FALSE, warning=FALSE, paged.print=FALSE}
library(geoR)
data(gambia)
```

```{r 09-Spatial-modeling-of-geostatistical-data-2}
head(gambia)
```

```{r 09-Spatial-modeling-of-geostatistical-data-3}
dim(gambia)
```

The analysis is done at the village level by aggregating the malaria tests by village.

```{r 09-Spatial-modeling-of-geostatistical-data-4, message=FALSE, warning=FALSE, paged.print=FALSE}
library(tidyverse)

gambia %>%
  count(x,y) %>%
  head()
```

### Prevalence

Grouping coordinates found 65 locations, or villages were the test were conducted.

```{r 09-Spatial-modeling-of-geostatistical-data-5}
gambia %>%
  count(x,y) %>%
  dim()
```

```{r 09-Spatial-modeling-of-geostatistical-data-6}
d <- gambia %>%
  group_by(x,y) %>%
  summarise(total = n(),
            positive = sum(pos),
            prev = positive / total,.groups = "drop")
head(d)
```

### Transforming coordinates

```{r 09-Spatial-modeling-of-geostatistical-data-7, message=FALSE, warning=FALSE}
library(sp)
library(rgdal)
sps <- SpatialPoints(d[, c("x", "y")],
  proj4string = CRS("+proj=utm +zone=28")
)
spst <- spTransform(sps, CRS("+proj=longlat +datum=WGS84"))
```

```{r 09-Spatial-modeling-of-geostatistical-data-8}
library(sf)

sps <- d %>%
  st_as_sf(coords=c(1,2),
           crs = "+proj=utm +zone=28")

sps %>%head()
```

```{r 09-Spatial-modeling-of-geostatistical-data-9}
spst <- sps %>%
  st_transform(crs = "+proj=longlat +datum=WGS84")
spst %>%head()
```

```{r 09-Spatial-modeling-of-geostatistical-data-10}
d1 <- spst %>%
  st_coordinates() %>%
  bind_cols(d) %>%
  rename(long=X,lat=Y)

d1 %>% head()
```

### Mapping prevalence

```{r 09-Spatial-modeling-of-geostatistical-data-11, message=FALSE, warning=FALSE}
library(leaflet)
library(viridis)

pal <- colorBin("viridis", 
                bins = c(0, 0.25, 0.5, 0.75, 1))

d1 %>%
  leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircles(lng = ~long, lat = ~lat, 
             color = ~ pal(prev)) %>%
  addLegend("bottomright",
            pal = pal, values = ~prev,
            title = "Prev.") %>%
  addScaleBar(position = c("bottomleft"))
```

### Environmental covariates

Model malaria prevalence using a covariate that indicates the altitude in The Gambia:

```{r 09-Spatial-modeling-of-geostatistical-data-12, message=FALSE, warning=FALSE}
library(raster)
r <- getData(name = "alt", 
             country = "GMB", 
             path = "data",
             mask = TRUE)
r
```

```{r 09-Spatial-modeling-of-geostatistical-data-13}
pal <- colorNumeric("viridis", 
                    values(r),
                    na.color = "transparent")

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addRasterImage(r, 
                 colors = pal, 
                 opacity = 0.5) %>%
  addLegend("bottomright",
            pal = pal, 
            values = values(r),
            title = "Altitude"
            ) %>%
  addScaleBar(position = c("bottomleft"))
```

Add the altitude:

```{r 09-Spatial-modeling-of-geostatistical-data-14}
# ?raster::extract

d2 <- d1 %>%
  mutate(alt = raster::extract(r,
                               d1[, c("long","lat")]))

d2 %>% head()
```

## Modeling

Steps to fit the model using the **SPDE approach** and the **R-INLA package**:

$$Y_i|P(x_i)\sim\text{Binomial}(N_i,P(x_i))$$ $$logit(P(x_i))=\beta_0+\beta_1 \text{x altitude + }S(x_i)$$

$$Cov(S(x_i),S(x_j))$$ $K_v(.)$ is the modified **Bessel function**. $v$ is the smoother parameter. $\rho=\sqrt{8v}/k$ is the **distance** where the spatial correlation is close to 0.1.

### Mesh construction

Build a **triangulated mesh** that covers The Gambia:

```{r 09-Spatial-modeling-of-geostatistical-data-15, message=FALSE, warning=FALSE, paged.print=FALSE}
library(INLA)
coo <- cbind(d2$long, d2$lat)


mesh <- inla.mesh.2d(
  loc = coo, 
  max.edge = c(0.1, 5),
  cutoff = 0.01)
```

Number of vertices:

```{r 09-Spatial-modeling-of-geostatistical-data-16}
mesh$n
```

```{r 09-Spatial-modeling-of-geostatistical-data-17}
plot(mesh)
points(coo, col = "red")
```

### Building the SPDE model on the mesh

The smoother parameter for the **stochastic partial differential equation SPDE** is $\alpha=v+d/2$ defined by $v=1$ and $d=2$:

    ?inla.spde2.matern

```{r 09-Spatial-modeling-of-geostatistical-data-18}
spde <- inla.spde2.matern(mesh = mesh, 
                          alpha = 2, 
                          constr = TRUE)
```

### Index set

Index set for the **SPDE model** where $s$ is the effect:

```{r 09-Spatial-modeling-of-geostatistical-data-19}
indexs <- inla.spde.make.index("s", spde$n.spde)
lengths(indexs)
```

### Projection matrix

```{r 09-Spatial-modeling-of-geostatistical-data-20}
A <- inla.spde.make.A(mesh = mesh, loc = coo)
```

### Prediction data

Specify the locations where to predict the prevalence:

```{r 09-Spatial-modeling-of-geostatistical-data-21}
dp <- raster::rasterToPoints(r)
dim(dp)
```

```{r 09-Spatial-modeling-of-geostatistical-data-22}
ra <- raster::aggregate(r, fact = 5, fun = mean)

dp <- raster::rasterToPoints(ra)
dim(dp)
```

```{r 09-Spatial-modeling-of-geostatistical-data-23}
coop <- dp[, c("x", "y")]
```

```{r 09-Spatial-modeling-of-geostatistical-data-24}
Ap <- inla.spde.make.A(mesh = mesh, loc = coop)
```

### Stack with data for estimation and prediction

Construct a **stack for estimation** called `stk.e` and a **stack for prediction** `stk.p`, then put them together in a **full stack** `stk.full`:

```{r 09-Spatial-modeling-of-geostatistical-data-25}
# stack for estimation stk.e
stk.e <- inla.stack(
  tag = "est",
  data = list(y = d2$positive, 
              numtrials = d2$total),
  A = list(1, A),
  effects = list(data.frame(b0 = 1, 
                            altitude = d2$alt), 
                 s = indexs))

# stack for prediction stk.p
stk.p <- inla.stack(
  tag = "pred",
  data = list(y = NA, 
              numtrials = NA),
  A = list(1, Ap),
  effects = list(data.frame(b0 = 1, 
                            altitude = dp[, 3]),
                 s = indexs))

# stk.full has stk.e and stk.p
stk.full <- inla.stack(stk.e, stk.p)
```

### Model formula

```{r 09-Spatial-modeling-of-geostatistical-data-26}
formula <- y ~ 0 + b0 + altitude + f(s, model = spde)
```

    ?inla()
    ?control.predictor

```{r 09-Spatial-modeling-of-geostatistical-data-27}
res <- inla(formula,
  family = "binomial", 
  Ntrials = numtrials,
  control.family = list(link = "logit"),
  data = inla.stack.data(stk.full),
  control.predictor = list(compute = TRUE, 
                           link = 1,
                           A = inla.stack.A(stk.full)))
```

```{r 09-Spatial-modeling-of-geostatistical-data-28}
summary(res)
```

## Mapping malaria prevalence

```{r 09-Spatial-modeling-of-geostatistical-data-29}
index <- inla.stack.index(stack = stk.full, 
                          tag = "pred")$data
```


```{r 09-Spatial-modeling-of-geostatistical-data-30}
# res$summary.fitted.values
```

```{r 09-Spatial-modeling-of-geostatistical-data-31}
prev_mean <- res$summary.fitted.values[index, "mean"]
prev_ll <- res$summary.fitted.values[index, "0.025quant"]
prev_ul <- res$summary.fitted.values[index, "0.975quant"]
```

```{r 09-Spatial-modeling-of-geostatistical-data-32}
pal <- colorNumeric("viridis", c(0, 1), na.color = "transparent")

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircles(
    lng = coop[, 1], lat = coop[, 2],
    color = pal(prev_mean)
  ) %>%
  addLegend("bottomright",
    pal = pal, values = prev_mean,
    title = "Prev."
  ) %>%
  addScaleBar(position = c("bottomleft"))
```

```{r 09-Spatial-modeling-of-geostatistical-data-33}
r_prev_mean <- rasterize(
  x = coop, y = ra, field = prev_mean,
  fun = mean
)
```

```{r 09-Spatial-modeling-of-geostatistical-data-34}
pal <- colorNumeric("viridis", c(0, 1), na.color = "transparent")

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addRasterImage(r_prev_mean, colors = pal, opacity = 0.5) %>%
  addLegend("bottomright",
    pal = pal,
    values = values(r_prev_mean), title = "Prev."
  ) %>%
  addScaleBar(position = c("bottomleft"))
```

```{r 09-Spatial-modeling-of-geostatistical-data-35}
r_prev_ll <- rasterize(
  x = coop, y = ra, 
  field = prev_ll,
  fun = mean
)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addRasterImage(r_prev_ll, colors = pal, opacity = 0.5) %>%
  addLegend("bottomright",
    pal = pal,
    values = values(r_prev_ll), title = "LL"
  ) %>%
  addScaleBar(position = c("bottomleft"))
```

```{r 09-Spatial-modeling-of-geostatistical-data-36}
r_prev_ul <- rasterize(
  x = coop, y = ra, field = prev_ul,
  fun = mean
)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addRasterImage(r_prev_ul, colors = pal, opacity = 0.5) %>%
  addLegend("bottomright",
    pal = pal,
    values = values(r_prev_ul), title = "UL"
  ) %>%
  addScaleBar(position = c("bottomleft"))
```

## Mapping exceedance probabilities

Calculate the **exceedance probabilities of malaria prevalence** being greater than a given threshold value.

$$P(p_1>c)=1-P(p_i\leq c)$$

Once the the marginals are found the **probability that malaria prevalence exceeds 20%** at this location is 0.6931.

How to calculate the marginals?

The **marginal distribution of the predictions** listed in the `res$marginals.fitted.values[index]` are indices of the **stk.full**:

```{r 09-Spatial-modeling-of-geostatistical-data-37}
index <- inla.stack.index(stack = stk.full, 
                          tag = "pred")$data
```

```{r 09-Spatial-modeling-of-geostatistical-data-38}
res2 <- inla(formula,
  family = "binomial", 
  Ntrials = numtrials,
  control.family = list(link = "logit"),
  data = inla.stack.data(stk.full),
  control.predictor = list(compute = TRUE, 
                           link = 1,
                           A = inla.stack.A(stk.full)),
  control.compute=list(return.marginals.predictor=TRUE))
```

```{r 09-Spatial-modeling-of-geostatistical-data-39}
summary(res2)
```

```{r 09-Spatial-modeling-of-geostatistical-data-40}
# res2$marginals.fitted.values[index]
```

```{r 09-Spatial-modeling-of-geostatistical-data-41, eval=FALSE}
marg <- res2$marginals.fitted.values[index][[1]]

1 - inla.pmarginal(q = 0.20, marginal = marg)
```

```{r 09-Spatial-modeling-of-geostatistical-data-42}
excprob <- sapply(res2$marginals.fitted.values[index],
                  FUN = function(marg){
                    1-inla.pmarginal(q = 0.20, 
                                     marginal = marg)})

head(excprob)
```

```{r 09-Spatial-modeling-of-geostatistical-data-43, eval=FALSE}
r_excprob <- rasterize(
  x = coop, y = ra, 
  field = excprob,
  fun = mean)
```

```{r 09-Spatial-modeling-of-geostatistical-data-44, eval=FALSE}
pal <- colorNumeric("viridis", c(0, 1), na.color = "transparent")

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addRasterImage(r_excprob, colors = pal, opacity = 0.5) %>%
  addLegend("bottomright",
    pal = pal,
    values = values(r_excprob), title = "P(p>0.2)"
  ) %>%
  addScaleBar(position = c("bottomleft"))
```

**END**

## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>

<summary>Meeting chat log</summary>

    LOG

</details>
