# Spatio-temporal modeling of geostatistical data. Air pollution in Spain

**Learning objectives:**

- Same workflow but this time with penalised complexity priors  
    - `inla.spde2pcmatern()`  
- Adding time in the analysis  
    
## Map

### Getting Spain

```{r mapch10, message=FALSE}
#spatial
library(lwgeom)
library(raster)
library(sf)
#data wrangling
library(dplyr)
#INLA
library(INLA)
#data viz
library(ggplot2)
library(viridis)

m <- getData(name = "GADM", country = "Spain", level = 0) # warning indicating that {geodata} should used
plot(m)
```

### Filter to only get main territory

```{r filterch10}
m <- m %>%
  st_as_sf() %>%
  st_cast("POLYGON") %>%
  mutate(area = st_area(.)) %>%
  arrange(desc(area)) %>%
  slice(1)
```


```{r transformspain}
m <- st_transform(m, 25830) 
ggplot(m) + 
  geom_sf() + 
  theme_bw() + 
  coord_sf(datum = st_crs(m)) # did not know that !
```

## Data!

### Getting it

We get them from [European Environment Agency](http://aidef.apps.eea.europa.eu/?source=%7B%22query%22%3A%7B%22match_all%22%3A%7B%7D%7D%2C%22display_type%22%3A%22tabular%22%7D)

We are using annual averages of $PM_2.5$ recorded at monitoring stations (2015, 2016,2017).

```{r gettingdata}
my_colnames <- c("ID", "CountryOrTerritory",	"ReportingYear",	"UpdateTime",
                 "StationLocalId",	"SamplingPointLocalId",	"SamplingPoint_Latitude",
                 "SamplingPoint_Longitude",	"Pollutant",	"AggregationType",
                 "Namespace",	"Unit",	"BeginPosition",	"EndPosition",
                 "Validity",	"Verification",	"DataCoverage",	"DataCapture",
                 "TimeCoverage",	"AQValue")


d <- read.csv("https://www.paulamoraga.com/assets/datasets/dataPM25.csv", 
              skip = 1, 
              header = FALSE,
              col.names = my_colnames)


```

### Cleaning it

Data are in 4326 (lat/long)

```{r selectingdata}
d <- d[, c(
  "ReportingYear", "StationLocalId",
  "SamplingPoint_Longitude",
  "SamplingPoint_Latitude",
  "AQValue"
)]
names(d) <- c("year", "id", "long", "lat", "value")
```


```{r}
p <- st_as_sf(data.frame(long = d$long, lat = d$lat),
              coords = c("long", "lat"))
st_crs(p) <- st_crs(4326)
p <- st_transform(p, 25830)
d[, c("x", "y")] <- st_coordinates(p) # new coords in 25830

ind <- st_intersects(m, p) # return a sparse geometry binary predicate 
d <- d[ind[[1]], ]  # it is a list with just "spain" so we simplify [[]] it
```

We plot it!

```{r}
ggplot(m) + geom_sf() + coord_sf(datum = NA) +
  geom_point(
    data = d, aes(x = x, y = y, color = value),
    size = 2
  ) +
  labs(x = "", y = "") +
  scale_color_viridis() +
  facet_wrap(~year) +
  theme_bw()
```

## Modeling!

$Y_{it}$ : $PM_{2.5}$ $i$ are the locations at $t$ times.

$$Y_{it} \sim N(\mu_{it}, \sigma^2_e )$$

$$\mu_{it} = \beta_0  + \xi(\boldsymbol{x}_i, t)$$

$\sigma_2²$ = variance of the measurement error (not correlated in space and time)

$\beta_0$ = intercept

$\xi(\boldsymbol{x}_i, t)$: random effect changes in time first order autoregressive dynamics and spatially correlated innovations


$$\xi(\boldsymbol{x}_i, t) = a \xi(\boldsymbol{x}_i, t-1) + w(\boldsymbol{x}_i, t) $$

First part is first order autoregressive process (think $Y_t = \beta_0 + \beta_1 Y_{t-1} + u_t$) and second part is the classic : 

> follows a zero-mean Gaussian distribution temporally independent but spatially dependent at each time with Matérn covariance function.

This is a "simple": we could imagine relation between time and space and had more covariate (temperature, precipitation, etc ...)

### Mesh construction

This is like the previous chapter: 

```{r mesh10}

coo <- cbind(d$x, d$y)
bnd <- inla.nonconvex.hull(st_coordinates(m)[, 1:2])
mesh <- inla.mesh.2d(
  loc = coo, boundary = bnd,
  max.edge = c(100000, 200000), # huge!
  cutoff = 1000
)

plot(mesh)
points(coo, col = "red")
```

### SPDE model on the mesh with penalised complexity

Picking prior for range and sigma:

```{r spde2pcmatern}
spde <- inla.spde2.pcmatern(
  mesh = mesh, alpha = 2, constr = TRUE, # same as last week
  prior.range = c(10000, 0.01), # P(range < 10000) = 0.01 / P(range < v_r) = p_r
  prior.sigma = c(3, 0.01) # P(sigma > 3) = 0.01 / P(sigma > v_s) = p_s
)
```

### Index

```{r index10}
timesn <- length(unique(d$year))
indexs <- inla.spde.make.index("s",
  n.spde = spde$n.spde,
  n.group = timesn
)
lengths(indexs) # pay attention to lengthS()
```

### projection matrix:

```{r projmatrix10}
group <- d$year - min(d$year) + 1 # we are just creating group similar to index
A <- inla.spde.make.A(mesh = mesh, loc = coo, 
                      group = group) # new stuff
# 299 x 2142 sparse Matrix of class "dgCMatrix"
```

### Grid:

```{r grid10}
bb <- st_bbox(m)
x <- seq(bb$xmin - 1, bb$xmax + 1, length.out = 50) # do not know if i should convert first or not ?
y <- seq(bb$ymin - 1, bb$ymax + 1, length.out = 50)
dp <- as.matrix(expand.grid(x, y)) 

p <- st_as_sf(data.frame(x = dp[, 1], y = dp[, 2]),
  coords = c("x", "y")
) # this could be a function no ?

st_crs(p) <- st_crs(25830)
ind <- st_intersects(m, p)
dp <- dp[ind[[1]], ]
plot(dp, asp = 1)
```

We just repeat it 3 times (3 years of data)

```{r}
dp <- rbind(cbind(dp, 1), cbind(dp, 2), cbind(dp, 3)) # 1,2,3 are recycled
dim(dp)
```

### Ap + stack

```{r ap10}
coop <- dp[, 1:2]
groupp <- dp[, 3]
Ap <- inla.spde.make.A(mesh = mesh, loc = coop, group = groupp)

stk.e <- inla.stack(
  tag = "est",
  data = list(y = d$value),
  A = list(1, A),
  effects = list(data.frame(b0 = rep(1, nrow(d))), s = indexs)
)

stk.p <- inla.stack(
  tag = "pred",
  data = list(y = NA),
  A = list(1, Ap),
  effects = list(data.frame(b0 = rep(1, nrow(dp))), s = indexs)
)

stk.full <- inla.stack(stk.e, stk.p)
```


### Model formula:

This will take a bit of time (around 1 min)! 

```{r modelformulainla}
rprior <- list(theta = list(prior = "pccor1", param = c(0, 0.9)))

formula <- y ~ 0 + b0 + f(s, #space time
  model = spde, 
  group = s.group, # years 
  control.group = list(model = "ar1", # auto regresive first order
                       hyper = rprior)
)


res <- inla(formula,
  data = inla.stack.data(stk.full),
  control.predictor = list(
    compute = TRUE,
    A = inla.stack.A(stk.full)
  )
)
```


I have not reproduce the summary of `res` but we have a range around 17857 m (from 14510 to 21721). 

```{r}
list_marginals <- list(
"b0" = res$marginals.fixed$b0,
"precision Gaussian obs" =
res$marginals.hyperpar$"Precision for the Gaussian observations",
"range" = res$marginals.hyperpar$"Range for s", # yes this is how they are called inside res
"stdev" = res$marginals.hyperpar$"Stdev for s",
"rho" = res$marginals.hyperpar$"GroupRho for s"
)


marginals <- data.frame(do.call(rbind, list_marginals))
marginals$parameter <- rep(names(list_marginals),
  times = sapply(list_marginals, nrow))
ggplot(marginals, aes(x = x, y = y)) + geom_line() +
  facet_wrap(~parameter, scales = "free") +
  labs(x = "", y = "Density") + theme_bw()

```

If I am correct rho (and GroupRho) are correlation between group (ie years)

## Mapping

We can do for every year we have data: 

```{r}
index <- inla.stack.index(stack = stk.full, tag = "pred")$data

dp <- data.frame(dp)
names(dp) <- c("x", "y", "time") # time is new

dp$pred_mean <- res$summary.fitted.values[index, "mean"]
dp$pred_ll <- res$summary.fitted.values[index, "0.025quant"]
dp$pred_ul <- res$summary.fitted.values[index, "0.975quant"]

library(reshape2) # I was to lazy
dpm <- melt(dp,
  id.vars = c("x", "y", "time"),
  measure.vars = c("pred_mean", "pred_ll", "pred_ul")
)

ggplot(m) + geom_sf() + coord_sf(datum = NA) +
  geom_tile(data = dpm, aes(x = x, y = y, fill = value)) +
  labs(x = "", y = "") +
  facet_wrap(variable ~ time) +
  scale_fill_viridis("PM2.5") +
  theme_bw()
```

## More on space time model:

Here: https://becarioprecario.bitbucket.io/inla-gitbook/ch-temporal.html

## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>
<summary> Meeting chat log </summary>

```
LOG
```
</details>
